---
output: github_document
always_allow_html: yes
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

# `val.meter`

<!-- badges: start -->
[![R-CMD-check](https://github.com/pharmaR/val.meter/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/pharmaR/val.meter/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/pharmaR/val.meter/graph/badge.svg)](https://app.codecov.io/gh/pharmaR/val.meter)
<!-- badges: end -->

_validation-ready package quantification supporting regulatory use of R_

```{r, include = FALSE, cache = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README",
  out.width = "100%",
  out.lines = 8,
  cache = TRUE
)

Sys.setenv(ASCIICAST_KNITR_SVG = "TRUE")
asciicast::init_knitr_engine(
  startup = quote(library(val.meter)),
  echo = TRUE,
  echo_input = FALSE
)
```

## Installation

```{r, eval = FALSE}
# install.packages("pak")
pak::pak("pharmaR/val.meter")
```

## Overview

Easily measure characteristics of packages, built to support the process of
_validation_ required in regulated industries, but generally useful to anyone
who wants to survey packages.

```{asciicast overview-metrics}
rpkg <- random_pkg(permissions = TRUE)
metrics(rpkg)
```

## Features

### Exploring metrics

`val.meter` makes it easy to find a listing of metrics and details of what those
metrics represent.

```{asciicast metrics, warning = FALSE}
metrics()
```

`metrics` actually represent a small set of the internally calculated package
data, which may go through a few stages of calculation before resulting in a
quantifiable metric. If you want to see _all_ the internal information, you can
explore the whole set of package data.

`metrics` are unique in that their data is required to be consistent and simple
(atomic) data.

```{asciicast metrics-all}
# see all implemented _data_ (a superset of metrics)
metrics(all = TRUE)
```

### Creating package objects

To start evaluating metrics, start by create a `pkg` (package) object. Package
objects represent a set of resources from which metadata is pulled and a
collection of package metadata. When we pass a `character` path or package name,
`val.meter` will search for the package from permitted sources.

```{asciicast pkg}
# calculate some data, using default conservative permissions
pkg("../val.meter")
```

We can also be declarative about exactly how we want to source our package
metadata.

```{asciicast pkg-explicit}
library(S7)

# initialize package with only our installed package as a resource
resrc <- convert("../val.meter", source_code_resource)
pkg(resrc)
```

#### Calculating metrics

Once we have declared _what_ we want to assess, we can then start calculating
metrics.

```{asciicast metrics-from-pkg}
p <- pkg("../val.meter")
metrics(p)
```

In this case, you'll notice that some metric calculations raised errors during
execution. Here you'll notice that we have not granted `val.meter` permission
to execute code or fetch data from APIs over the network. 

We can also access individual metrics by accessing our `pkg` object like a list.

```{asciicast metric-dep-count}
p$dependency_count
```

Packages are _lazy_ :zzz:! Since some metrics can be computationally intensive,
we only calculate them as their needed. After they're calculated the first
time, the result is saved so you can access it whenever you need it.

#### Managing execution permissions

We can opt-in to more extensive capabilities by giving our package more
permissive execution. We pass `permissions(TRUE)` to grant blanket permission
to all capabilities.

```{asciicast metrics-permissive, error = TRUE}
p <- pkg("../val.meter", permissions = "network")
metrics(p)
```

And digging a bit deeper, we'll see that `r_cmd_check` itself requires
additional permissions to grant `val.meter` the ability to execute code on your
machine.

```{asciicast metric-r-cmd-check-error, error = TRUE}
p$r_cmd_check
```
